\maketitle

\frontmatter

\section*{Preface}

At the end of 2019 I had started my first real job and for the first time in my life started making real money. As a 19 year old living at home with no meaningful expenses I quickly learned about the common vices of our world, alcohol and gambling. My beverage of choice was a single malt scotch whiskey, in particular Loch Lomond and Laphroaig, and my favourite method of gambling was on the stock market. On the night of the 2020 US election I wasn't celebrating or commiserating my political party, I was trading 500x leveraged CFDs on NSDQ100 and SPX500 and took my account from about 200USD to about 1800USD. I wasn't a complete fool, I withdrew enough money that I had completely covered what I had put in, so I was playing with the houses money, but, I had gained something else that night. Dangerous amounts of misplaced confidence. In the next few weeks I lost all of my winnings, then, started funding my account to keep playing until one night I realised I'd gambled away my entire last weeks pay. The anxiety of watching the number go down was crippling, and really made the lowest point in my life a lot lower. I cleaned myself up and got my shit together but that part of my story is not what got me here, no. I vowed revenge. 

Every failure is a learning opportunity, and there were a lot of lessons learned. One, know your risk tolerance. You will lose when you lose your nerve and make decisions based on panic. Two, I don't have the nerve for day trading.

I followed lesson one, instead of gambling big on speculative mining companies and making 2-3x in a few weeks, instead, I put my money into ETFs and made respectable returns. Two I decided to give up my dream of being a day trader... and turned it into a dream of having a program to do it for me.

In 2021 I started university. In my second semester, MZB126, we learned about ODEs and it changed my world. I felt like I had just learned the why of all of the mathematics I had ever sat through and had the tools to understand everything, I felt like I was staring into the face of god themself. I wrote a physics engine in MATLAB since I had learned that in the same class, and made a lil rocket game where you flew around a rocket. I decided in this time too that computational physics is my future and just coded. I went through maybe 5 versions of that rocket game, made a 2D projection of an n-cube you could move around and play with and I learned how much you can learn from doing these sorts of self-motivated projects. I realised that if I practiced something, I would learn, it and become good at it.

Shortly after Christmas 2021 the rocket and the cubes were reaching a fairly polished state and it was time for a new project, I already knew what it would be. I had one learning objective and one goal; learn Python my second programming language, and, make a trading algorithm to take my revenge. I would start CBG (Crystal Ball Gazing).

I knew that I was far too emotional to day-trade. The simple fact is that If I felt something I would find technical indicators to support the decision I had already made, and then lose money. My goal was to create a program which would look at the market and decide an entry, stop-loss, and take-profit price and execute the trades on etoro. No emotions, only silicon and mathematics. I learned a lot in that first attempt, but frankly, the amount that I did not know meant that that first attempt was nothing more than a learning experience. My next major attempt came at the end of 2022 a year later when I did CAB420 Machine Learning, I implemented all sorts of models, but, I never-ever created a program I believed in.

There have been plenty of minor attempts after this whenever I had an idea or learned something new, but doing my final year of physics, then my honours of mathematics I simply did not have the time or mental capacity to tackle this project. 

Until now.

After finishing my Honours I decided to look for work, and as a part of looking for work I decided that having some of my projects on GitHub would hopefully supplement my resum\'e neatly. Since I am looking for analytical roles including finance I decide that I should dig this project up for the world to see. I think I still have every major version of CBG, but my old code disgusts me, so, I essentially restarted from scratch.

And that's where we are today, that is what this is. Part complement to my resum\'e, part revenge for that money I lost.

The idea of this document is serve as a tour guide for the controlled chaos happening elsewhere. I am hoping to write down the ideas and mathematical derivations that underpin what I am doing, so hopefully another person could actually understand wtf is going on here. And maybe, just maybe, convince someone that I would be a valuable member of their team.

\tableofcontents